Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson's \emph{Irises}

### 2020-09-13 penguins.Rmd
minor intro updates

In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three Iris species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935]. The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning. Anderson's *Iris* dataset is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the data. As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics, computational methods, and data science courses and materials.

There are several reasons that practitioners and educators may want to retire Anderson's *Iris* dataset in modern data science materials, including:

- Anderson's  *Iris* dataset entirely lacks metadata, limiting meaningful interpretation and discussion of research methods, analyses, and outcomes
- Two of its five variables, `Sepal.Width` and `Sepal.Length`, are not intuitive to most users
- In 1936, the *Iris* data were published by eugenicist R.A. Fisher in *Annals of Eugenics*, and are now inextricably linked to Fisher's eugenics work; in fact, the dataset is commonly referred to as *Fisher's Iris data*
- The *Iris* dataset contains only complete cases (i.e. no missing values) and equal sample sizes (*n* = 50 for each *Iris* species) - more perfect than real-world datasets learners are likely to encounter "in the wild"
- The single categorical variable (`Species`) limits options for analyses

@@

Given these motivations to retire *Iris*, we sought to identify, then make readily available, a suitable alternative dataset. Considering *Iris* usage in data science and statistics materials, we established the following criteria for a suitable alternative:

>>>>

In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three Iris species: *Iris setosa*, *Iris versicolor*, and *Iris virginica* [@anderson_irises_1935]. The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of the data, including linear relationships and multivariate normality, make it amenable for introducing a wide range of statistical methods including data wrangling, visualization, regression, multivariate exploration, and machine learning. Anderson's *Iris* dataset is built into a number of software packages including the auto-installed datasets package in R [@r_core_team_r_2019], Python's scikit-learn machine learning library [@pedregosa_scikit-learn_2011], and the SAS Sashelp library (SAS Institute, Cary NC), which have facilitated widespread use of the data. As a result, eighty-five years after the data were initially published, Anderson’s *Iris* measurements are ubiquitous in statistics, computational methods, and data science courses and materials.

There are a number of reasons that practitioners and educators may want to retire Anderson's *Iris* dataset in their modern data science materials. Anderson's *Iris* dataset lacks metadata almost entirely [@anderson_irises_1935], limiting meaningful interpretation and discussion of research methods, analyses, and outcomes. Of five variables in the dataset, two (`Sepal.Width` and `Sepal.Length`) are unintuitive for most non-botanists, and even with explanation the difference between *Iris* *petal* and *sepal* dimensions is not obvious. The *Iris* dataset contains only complete cases (there are no missing values) and equal sample size for all 3 species (*n* = 50 for all), which is cleaner than most real-world datasets learners are likely to encounter "in the wild." The single factor (`Species`) in the *Iris* data limits options for analyses. Finally, due to their publication in (and inextricable connection to) eugenicist and statistician R.A. Fisher's 1936 paper in *Annals of Eugenics* [@fisher_use_1936], the *Iris* dataset is burdened by a history in racist research.

@@

Given these motivations to retire *Iris*, we sought to identify, then make readily available, a suitable alternative dataset. Considering the positive and negative characteristics of the *Iris* dataset in data science and statistics materials, we established the following criteria for a suitable alternative:

### 2020-09-13 penguins.tex
minor intro updates

In 1935, American botanist Edgar Anderson measured petal and sepal
structural dimensions (length and width) for 50 flowers from three Iris
species: \emph{Iris setosa}, \emph{Iris versicolor}, and \emph{Iris
virginica} \citep{anderson_irises_1935}. The manageable but non-trivial
size (5 variables and 150 total observations) and characteristics of the
data make it amenable for introducing a wide range of statistical
methods including data wrangling, visualization, regression,
multivariate exploration, and machine learning. Anderson's \emph{Iris}
dataset is built into a number of software packages including the
auto-installed datasets package in R \citep{r_core_team_r_2019},
Python's scikit-learn machine learning library
\citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library (SAS
Institute, Cary NC), which have facilitated widespread use of the data.
As a result, eighty-five years after the data were initially published,
Anderson's \emph{Iris} measurements are ubiquitous in statistics,
computational methods, and data science courses and materials.

There are several reasons that practitioners and educators may want to
retire Anderson's \emph{Iris} dataset in modern data science materials,
including:

\begin{itemize}
\tightlist
\item
  Anderson's \emph{Iris} dataset entirely lacks metadata, limiting
  meaningful interpretation and discussion of research methods,
  analyses, and outcomes
\item
  Two of its five variables, \texttt{Sepal.Width} and
  \texttt{Sepal.Length}, are not intuitive to most users
\item
  In 1936, the \emph{Iris} data were published by eugenicist R.A. Fisher
  in \emph{Annals of Eugenics}, and are now inextricably linked to
  Fisher's eugenics work; in fact, the dataset is commonly referred to
  as \emph{Fisher's Iris data}
\item
  The \emph{Iris} dataset contains only complete cases (i.e.~no missing
  values) and equal sample sizes (\emph{n} = 50 for each \emph{Iris}
  species) - more perfect than real-world datasets learners are likely
  to encounter ``in the wild''
\item
  The single categorical variable (\texttt{Species}) limits options for
  analyses
\end{itemize}

Given these motivations to retire \emph{Iris}, we sought to identify,
then make readily available, a suitable alternative dataset. Considering
\emph{Iris} usage in data science and statistics materials, we
established the following criteria for a suitable alternative:

@@

\caption{\label{tab:unnamed-chunk-3}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
\centering
\begin{tabular}[t]{c|c|c|c|c|c|c|c}
\hline
\multicolumn{4}{c|}{Penguins cluster assignments} & \multicolumn{4}{c}{Iris cluster assignments} \\
\cline{1-4} \cline{5-8}
Cluster & Adélie & chinstrap & gentoo & Cluster & setosa & versicolor & virginica\\
\hline
1 & 4 & 54 & 6 & 1 & 50 & NA & NA\\
\hline
2 & 147 & 5 & 1 & 2 & NA & 48 & 4\\
\hline
3 & NA & 9 & 116 & 3 & NA & 2 & 46\\
\hline
\end{tabular}
\end{table}

>>>>

In 1935, American botanist Edgar Anderson measured petal and sepal
structural dimensions (length and width) for 50 flowers from three Iris
species: \emph{Iris setosa}, \emph{Iris versicolor}, and \emph{Iris
virginica} \citep{anderson_irises_1935}. The manageable but non-trivial
size (5 variables and 150 total observations) and characteristics of the
data, including linear relationships and multivariate normality, make it
amenable for introducing a wide range of statistical methods including
data wrangling, visualization, regression, multivariate exploration, and
machine learning. Anderson's \emph{Iris} dataset is built into a number
of software packages including the auto-installed datasets package in R
\citep{r_core_team_r_2019}, Python's scikit-learn machine learning
library \citep{pedregosa_scikit-learn_2011}, and the SAS Sashelp library
(SAS Institute, Cary NC), which have facilitated widespread use of the
data. As a result, eighty-five years after the data were initially
published, Anderson's \emph{Iris} measurements are ubiquitous in
statistics, computational methods, and data science courses and
materials.

There are a number of reasons that practitioners and educators may want
to retire Anderson's \emph{Iris} dataset in their modern data science
materials. Anderson's \emph{Iris} dataset lacks metadata almost entirely
\citep{anderson_irises_1935}, limiting meaningful interpretation and
discussion of research methods, analyses, and outcomes. Of five
variables in the dataset, two (\texttt{Sepal.Width} and
\texttt{Sepal.Length}) are unintuitive for most non-botanists, and even
with explanation the difference between \emph{Iris} \emph{petal} and
\emph{sepal} dimensions is not obvious. The \emph{Iris} dataset contains
only complete cases (there are no missing values) and equal sample size
for all 3 species (\emph{n} = 50 for all), which is cleaner than most
real-world datasets learners are likely to encounter ``in the wild.''
The single factor (\texttt{Species}) in the \emph{Iris} data limits
options for analyses. Finally, due to their publication in (and
inextricable connection to) eugenicist and statistician R.A. Fisher's
1936 paper in \emph{Annals of Eugenics} \citep{fisher_use_1936}, the
\emph{Iris} dataset is burdened by a history in racist research.

Given these motivations to retire \emph{Iris}, we sought to identify,
then make readily available, a suitable alternative dataset. Considering
the positive and negative characteristics of the \emph{Iris} dataset in
data science and statistics materials, we established the following
criteria for a suitable alternative:

@@

\caption{\label{tab:unnamed-chunk-3}K-means cluster assignments by species based on penguin bill length (mm) and depth (mm), and iris petal length (cm) and width (cm).}
\centering
\begin{tabular}[t]{c|c|c|c|c|c|c|c}
\hline
\multicolumn{4}{c|}{Penguins cluster assignments} & \multicolumn{4}{c}{Iris cluster assignments} \\
\cline{1-4} \cline{5-8}
Cluster & Adélie & chinstrap & gentoo & Cluster & setosa & versicolor & virginica\\
\hline
1 & NA & 9 & 116 & 1 & NA & 48 & 4\\
\hline
2 & 4 & 54 & 6 & 2 & 50 & NA & NA\\
\hline
3 & 147 & 5 & 1 & 3 & NA & 2 & 46\\
\hline
\end{tabular}
\end{table}
